---
title: "Reproducible Research: Peer Assessment 1"
author: "Eric L"
date: "January 14, 2016"
output: 
  html_document:
    keep_md: true
---

## 1. Loading and preprocessing the data

### 1.1 Load the data

```{r}
activity <- read.csv("activity.csv")
head(activity); str(activity)
```

### 1.2 Pre-processing: convert date from character to date class

```{r}
activity$date <- as.Date(activity$date)
head(activity); str(activity)
```

## 2. What is mean total # of steps taken per day?  For this part ignore NAs

```{r}
## Subset data to exclude NAs
activitynona <- activity[!is.na(activity$steps),]
head(activitynona); str(activitynona)
```

### 2.1 Calculate the total # of steps per day
```{r}
stepsperday <- aggregate(steps ~ date, data = activitynona, FUN = "sum")
head(stepsperday); str(stepsperday)
```

### 2.2, 2.3 Create a histogram & summaries of total # of steps per day
```{r}
hist(stepsperday$step, main="Histogram of total # of steps per day",xlab="# of steps per day")
summary(stepsperday$steps)
```

## 3. What is the average daily pattern?

### 3.1 Make a time series plot of total # of steps by 5-minute interval
### 3.2 Which 5-minute interval on average across all days contains the max # of steps?
```{r}
stepsperinterval <- aggregate(steps ~ interval, data = activitynona, FUN = "mean")
head(stepsperinterval); str(stepsperinterval)
plot(stepsperinterval$interval, stepsperinterval$steps,type = "l",
     xlab = "interval", ylab = "mean # of steps per interval", main = "Average Daily Pattern")

summary(stepsperinterval$steps)
maxsteps <- max(stepsperinterval$steps)
maxinterval <- stepsperinterval[maxsteps == stepsperinterval$steps,]$interval

string <- paste("max # of steps on average is", signif(maxsteps,4), "at interval", maxinterval)
text(maxinterval,maxsteps + 4,string)
```

## 4. Imputing Missing Values

### 4.1.Calculate and report the total number of missing values in the dataset 
### (i.e. the total number of rows with NAs, which is same of missing step values)

```{r}
numna <- sum(is.na(activity$steps))
paste("The number of missing step values, i.e. rows is",numna)
```

### 4.2.Devise a strategy for filling in all of the missing values in the dataset. 

### Based on the variability seen in the plot of Average Daily Pattern.  Impute missing 
### values using the mean value for the interval over all days, already done above.

### 4.3.Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
## Rename steps variable to meansteps
library(plyr)
stepsperinterval <- rename(stepsperinterval, c("steps"="meansteps"))
head(stepsperinterval); str(stepsperinterval)

## Merge original activity data with meansteps data
activityrawsum <- merge(activity, stepsperinterval, by = "interval")
head(activityrawsum); str(activityrawsum)

## impute missing steps values using meansteps for that interval
activityrawsum$imputesteps <- activityrawsum$steps
my.na <- is.na(activityrawsum$steps)
activityrawsum$imputesteps[my.na] <-activityrawsum$meansteps[my.na]
head(activityrawsum); str(activityrawsum)
```

### 4.4 Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}
stepsperdayimputed <- aggregate(imputesteps ~ date, data = activityrawsum, FUN = "sum")
head(stepsperdayimputed); str(stepsperdayimputed)

hist(stepsperdayimputed$imputestep, main="Histogram of total # of steps (imputed) per day",
     xlab="# of steps (imputed) per day")
summary(stepsperdayimputed$imputesteps)
summary(stepsperday$steps)
```

### Summary stats mean and median are very similar for original & imputed step values

## 5. Are there differences in activity patterns between weekdays & weekends?

### 5.1.Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
activitynona$weekday <- weekdays(activitynona$date)
head(activitynona); str(activitynona)

## This function create a vector typeall with values weekend or weekday based on day
daytypenew <- function(weekday) {
  typeall <- character(0)
  for (i in 1:length(weekday)) {
    if (weekday[i] %in% c("Saturday","Sunday")) {
      type <- "weekend"
    } else {
      type <- "weekday"
    }
      typeall <- c(typeall,type)
  }
  typeall
}

activitynona$daytype <- daytypenew(activitynona$weekday)
head(activitynona); tail(activitynona); str(activitynona)
table(activitynona$daytype)
```

### 5.2.Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r}
stepsperinttype <- aggregate(steps ~ interval + daytype, data = activitynona, FUN = "mean")
head(stepsperinttype); str(stepsperinttype)

library(ggplot2)
qplot(interval, steps, data = activitynona, facets = daytype ~ ., geom = "line")
```


